---
title: "AppendixA"
date: "2024-02-21"
---

# Appendix A - Fixing Classification Error (Column: EVTYPE)

There are only 48 official classification for the storm data according to NOAA. However, after filtering out the columns we want, we still have about 440 unique classification. Massive clean up is required.

At first glance, a lot of the classification are merely spelling errors or in abbreviated form. However, some classification need to be explored further by examining the remarks column.

## Strategy on Clean Up EVTYPE Misclassification

We have the following methods to re-classified.

**Strategy 1**

The most ideal re-classification method is to read the remarks column of all those classification that are not under the official classification and determine the actual nature of storm type. However, this is a extremely tedious task. It might not be effective when our analysis try to account for major casualties and economic impact.

**Strategy 2**

The second strategy is to re-classified the storm event manually on the obvious error such as spelling error and abbreviations. We will target unique classification of less than 100.

**Conclusion**

After much consideration, we will adopt the second strategy as it is more cost effective in terms of time spent.

### Setting Environment

```{r, message=FALSE}
# Import Library
library(dplyr)
library(stringr)
library(ggplot2)
library(reshape2)
```

### Read File

```{r}
stormData <- read.csv("StormData.csv.bz2")
```

### Reduce data size

```{r}
reducedStormData <- filter(stormData, FATALITIES != 0 | INJURIES !=0 |PROPDMG !=0 | CROPDMG != 0)
reducedStormData <- select(reducedStormData, STATE, BGN_DATE, EVTYPE, FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP,REMARKS)
reducedStormData$EVTYPE <- trimws(reducedStormData$EVTYPE, which = c("both"), whitespace = "[ \t\r\n]")
reducedStormData$EVTYPE <- tolower(reducedStormData$EVTYPE)
reducedStormData$EVTYPE <- toupper(reducedStormData$EVTYPE)
reducedStormData$EVTYPE <- str_squish(reducedStormData$EVTYPE)
str(reducedStormData)
```

```{r}
evtypeLst <- sort(unique(reducedStormData$EVTYPE))
evtypeLst
```

**As shown above, we have 440 unique classification.**

```{r}
# Number of unique classification
numOldEvtypes <- length(evtypeLst)
```

### Re-Classified Obvious Error

```{r}
# Start Obvious Classification
reducedStormData$EVTYPE[grepl("ASTRONOMICAL HIGH TIDE", reducedStormData$EVTYPE)] <- "ASTRONOMICAL LOW TIDE"
reducedStormData$EVTYPE[grepl("AVALANCE", reducedStormData$EVTYPE)] <- "AVALANCHE"


# Blizzard
reducedStormData$EVTYPE[grepl("BLIZZARD/WINTER STORM", reducedStormData$EVTYPE)] <- "BLIZZARD"
reducedStormData$EVTYPE[grepl("GROUND BLIZZARD", reducedStormData$EVTYPE)] <- "BLIZZARD"


# Coastal Flood
reducedStormData$EVTYPE[grepl("^COASTAL*", reducedStormData$EVTYPE)] <- "COASTAL FLOOD"
reducedStormData$EVTYPE[grepl("EROSION/CSTL FLOOD", reducedStormData$EVTYPE)] <- "COASTAL FLOOD"


# COLD/WIND CHILL
reducedStormData$EVTYPE[grepl("^COLD*", reducedStormData$EVTYPE)] <- "COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("UNSEASONABLE COLD", reducedStormData$EVTYPE)] <- "COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("UNSEASONABLY COLD", reducedStormData$EVTYPE)] <- "COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("COOL AND WET", reducedStormData$EVTYPE)] <- "COLD/WIND CHILL"


# Dust Devil/Strom
reducedStormData$EVTYPE[grepl("^DUST DEVIL*", reducedStormData$EVTYPE)] <- "DUST DEVIL"
reducedStormData$EVTYPE[grepl("^DUST STORM*", reducedStormData$EVTYPE)] <- "DUST STORM"
reducedStormData$EVTYPE[grepl("^DRY MICROBURST*", reducedStormData$EVTYPE)] <- "DUST STORM"
reducedStormData$EVTYPE[grepl("BLOWING DUST", reducedStormData$EVTYPE)] <- "DUST STORM"


# Excessive Heat
reducedStormData$EVTYPE[grepl("^EXTREME HEAT*", reducedStormData$EVTYPE)] <- "EXCESSIVE HEAT"
reducedStormData$EVTYPE[grepl("^RECORD HEAT*", reducedStormData$EVTYPE)] <- "EXCESSIVE HEAT"
reducedStormData$EVTYPE[grepl("RECORD/EXCESSIVE HEAT", reducedStormData$EVTYPE)] <- "EXCESSIVE HEAT"


# Extreme Cold/Wind Chill
reducedStormData$EVTYPE[grepl("^EXTREME COLD*", reducedStormData$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("^EXTREME WIND*", reducedStormData$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("^EXTENDED*", reducedStormData$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("EXCESSIVE WETNESS", reducedStormData$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("RECORD COLD", reducedStormData$EVTYPE)] <- "EXTREME COLD/WIND CHILL"

# Flash Flood
reducedStormData$EVTYPE[grepl("^FLASH FLOOD*", reducedStormData$EVTYPE)] <- "FLASH FLOOD"

# Other Flood
reducedStormData$EVTYPE[grepl("^FLOOD*", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("^URBAN*", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("^RIVER.*FLOOD", reducedStormData$EVTYPE)] <- "FLOOD"

reducedStormData$EVTYPE[grepl("^MINOR FLOODING", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("MAJOR FLOOD", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("BREAKUP FLOODING", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("SMALL STREAM FLOOD", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("RURAL FLOOD", reducedStormData$EVTYPE)] <- "FLOOD"

# Freezing Fog
reducedStormData$EVTYPE[grepl("FOG", reducedStormData$EVTYPE)] <- "FREEZING FOG"

# Frost/Freeze
reducedStormData$EVTYPE[grepl("FROST\\FREEZE", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("FROST", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("FREEZE", reducedStormData$EVTYPE)] <- "FROST/FREEZE"

reducedStormData$EVTYPE[grepl("FREEZING RAIN", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("FREEZING DRIZZLE", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("FREEZING SPRAY", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("^HYPOTHERMIA.*", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("HYPERTHERMIA/EXPOSURE", reducedStormData$EVTYPE)] <- "FROST/FREEZE"


# Hail
reducedStormData$EVTYPE[grepl("^HAIL*", reducedStormData$EVTYPE)] <- "HAIL"


# Heat
reducedStormData$EVTYPE[grepl("^HEAT *", reducedStormData$EVTYPE)] <- "HEAT"


# Heavy Rain
reducedStormData$EVTYPE[grepl("^RAIN*", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("^HEAVY RAIN*", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("HEAVY PRECIPITATION", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("HEAVY SHOWER", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("EXCESSIVE RAINFALL", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("RECORD RAINFALL", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("UNSEASONAL RAIN", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("HVY RAIN", reducedStormData$EVTYPE)] <- "HEAVY RAIN"


# Heavy Snow
reducedStormData$EVTYPE[grepl("^HEAVY SNOW*", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("BLOWING SNOW", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("FALLING SNOW/ICE", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("LATE SEASON SNOW", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("EXCESSIVE SNOW", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("HEAVY LAKE SNOW", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("RECORD SNOW", reducedStormData$EVTYPE)] <- "HEAVY SNOW"


# High Surf
reducedStormData$EVTYPE[grepl("^HEAVY SURF*", reducedStormData$EVTYPE)] <- "HIGH SURF"
reducedStormData$EVTYPE[grepl("HAZARDOUS SURF", reducedStormData$EVTYPE)] <- "HIGH SURF"
reducedStormData$EVTYPE[grepl("HIGH SURF ADVISORY", reducedStormData$EVTYPE)] <- "HIGH SURF"
reducedStormData$EVTYPE[grepl("HIGH SWELLS", reducedStormData$EVTYPE)] <- "HIGH SURF"
reducedStormData$EVTYPE[grepl("HIGH WAVES", reducedStormData$EVTYPE)] <- "HIGH SURF"


# High Wind
reducedStormData$EVTYPE[grepl("^HIGH WIND*", reducedStormData$EVTYPE)] <- "HIGH WIND"


# Hurricane (Typhoon)
reducedStormData$EVTYPE[grepl("^HURRICANE*", reducedStormData$EVTYPE)] <- "HURRICANE (TYPHOON)"
reducedStormData$EVTYPE[grepl("^TYPHOON", reducedStormData$EVTYPE)] <- "HURRICANE (TYPHOON)"


# Ice Storm
reducedStormData$EVTYPE[grepl("^ICE*", reducedStormData$EVTYPE)] <- "ICE STORM"
reducedStormData$EVTYPE[grepl("BLACK ICE", reducedStormData$EVTYPE)] <- "ICE STORM"
reducedStormData$EVTYPE[grepl("^GLAZE", reducedStormData$EVTYPE)] <- "ICE STORM"


# Lakeshore Flood
reducedStormData$EVTYPE[grepl("^LAKE FLOOD", reducedStormData$EVTYPE)] <- "LAKESHORE FLOOD"


# Lake-Effect Snow
reducedStormData$EVTYPE[grepl("^LAKE EFFECT SNOW", reducedStormData$EVTYPE)] <- "LAKE-EFFECT SNOW"


# Lightning
reducedStormData$EVTYPE[grepl("^LIGHT*", reducedStormData$EVTYPE)] <- "LIGHTNING"
reducedStormData$EVTYPE[grepl("^LIGNTNING", reducedStormData$EVTYPE)] <- "LIGHTNING"


# Marine Thunderstorm Wind
reducedStormData$EVTYPE[grepl("^MARINE TSTM WIND", reducedStormData$EVTYPE)] <- "MARINE THUNDERSTORM WIND"


# Rip Current
reducedStormData$EVTYPE[grepl("^RIP CURRENT.*", reducedStormData$EVTYPE)] <- "RIP CURRENT"


# Strong Wind 
reducedStormData$EVTYPE[grepl("^GUST*", reducedStormData$EVTYPE)] <- "STRONG WIND"
reducedStormData$EVTYPE[grepl("^STRONG WIND*", reducedStormData$EVTYPE)] <- "STRONG WIND"
reducedStormData$EVTYPE[grepl("^WIND", reducedStormData$EVTYPE)] <- "STRONG WIND"


# Thunderstorm Wind
reducedStormData$EVTYPE[grepl("^TSTM*", reducedStormData$EVTYPE)] <- "THUNDERSTORM WIND"
reducedStormData$EVTYPE[grepl("^THU*", reducedStormData$EVTYPE)] <- "THUNDERSTORM WIND"
reducedStormData$EVTYPE[grepl("^TUNDERSTORM*", reducedStormData$EVTYPE)] <- "THUNDERSTORM WIND"
reducedStormData$EVTYPE[grepl("^SEVERE*", reducedStormData$EVTYPE)] <- "THUNDERSTORM WIND"


# Tornado
reducedStormData$EVTYPE[grepl("^TORNADO*", reducedStormData$EVTYPE)] <- "TORNADO"
reducedStormData$EVTYPE[grepl("^TORNDAO*", reducedStormData$EVTYPE)] <- "TORNADO"


# Tropical Storm
reducedStormData$EVTYPE[grepl("^TROPICAL STORM*", reducedStormData$EVTYPE)] <- "TROPICAL STORM"


# Waterspout
reducedStormData$EVTYPE[grepl("^WATERSPOUT*", reducedStormData$EVTYPE)] <- "WATERSPOUT"


# Winter Storm
reducedStormData$EVTYPE[grepl("^WINTER STORM*", reducedStormData$EVTYPE)] <- "WINTER STORM"


# Winter Weather
reducedStormData$EVTYPE[grepl("^WINTER WEATHER*", reducedStormData$EVTYPE)] <- "WINTER WEATHER"
reducedStormData$EVTYPE[grepl("WINTRY MIX", reducedStormData$EVTYPE)] <- "WINTER WEATHER"
reducedStormData$EVTYPE[grepl("^SNOW.*", reducedStormData$EVTYPE)] <- "WINTER WEATHER"



# Wildfire
reducedStormData$EVTYPE[grepl("^WILD.*FIRE*", reducedStormData$EVTYPE)] <- "WILDFIRE"
reducedStormData$EVTYPE[grepl("^BRUSH FIRE", reducedStormData$EVTYPE)] <- "WILDFIRE"
reducedStormData$EVTYPE[grepl("^FOREST FIRE*", reducedStormData$EVTYPE)] <- "WILDFIRE"
reducedStormData$EVTYPE[grepl("^GRASS FIRE*", reducedStormData$EVTYPE)] <- "WILDFIRE"


evtypeLst <- sort(unique(reducedStormData$EVTYPE))
print(paste("Before re-classification, we have unqiue class of:", as.character(numOldEvtypes)))
print(paste("After re-classification, we have unqiue class of:", as.character(length(evtypeLst))))


```

```{r}
evtypeLst <- sort(unique(reducedStormData$EVTYPE))
evtypeLst
```

## Checking Obvious Misclassification

### Checking on classification ?

```{r}
filter(reducedStormData, EVTYPE=="?")
```

**No data, no remarks. Hence row to be removed**.

```{r}
reducedStormData <- reducedStormData[reducedStormData$EVTYPE != "?",]
str(reducedStormData)
```

### Checking classification "APACHE COUNTY"

```{r}
ac <- filter(reducedStormData, EVTYPE=="APACHE COUNTY")
ac[,"REMARKS"]
```

**This is to reclassified to "THUNDERSTORM WIND"**

```{r}
# Reclassified APACHE COUNTY
reducedStormData$EVTYPE[grepl("APACHE COUNTY", reducedStormData$EVTYPE)] <- "THUNDERSTORM WIND"
```
