---
title: "Reproducible Research Project Final"
date: "2024-02-20"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of Human and Economic Impact of Storms in United States
This report and teh relevant data is available at [https://github.com/mlnotes2718/RepResearchFinalProject](https://github.com/mlnotes2718/RepResearchFinalProject)

## SYNOPSIS
The following is an analysis of various kind of storm is the United States since 1950s. This analysis focus on human casualties and financial impact causes by various kind of storm.


## DATA PROCESSING

### Setting Envionment
```{r, message=FALSE}
# Import Library
library(dplyr)
library(ggplot2)
library(reshape2)
```

### Data Source and Documentation

Data Source: [U.S. National Oceanic and Atmospheric Administration (NOAA) - Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

URL: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

Documentation of Data Source: 
[National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

[National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)



### Download Data File
```{r, cache = TRUE}
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileURL, destfile = "./StormData.csv.bz2")
```

### Data Dictionary
|Variable Name | Description |
|---|---|
|stormData | The raw data that is read from StormData.csv.bz2|
|reducedData | raw dataset with only only 7 columns |


### Read File
```{r, cache=TRUE}
stormData <- read.csv("StormData.csv.bz2")
```

File Structure
```{r}
str(stormData)
```

### Processing Data
1. First we will filter out those data that has fatalities, injuries, property damage and crop damage. 
```{r}
reducedStormData <- filter(stormData, FATALITIES != 0 | INJURIES !=0 |PROPDMG !=0 | CROPDMG != 0)
```

```{r}
reducedStormData <- select(reducedStormData, STATE, BGN_DATE, EVTYPE, FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP,REMARKS)
```

2. Check for NA
```{r}
stormTypeNA <- is.na(reducedStormData$EVTYPE)
filter(reducedStormData, stormTypeNA)
```
No NA for storm type


```{r}
casualtiesNA <- is.na(reducedStormData$FATALITIES | reducedStormData$INJURIES)
filter(reducedStormData, casualtiesNA)
```

No NA for fatalities or injuries.

```{r}
damageNA <- is.na(reducedStormData$PROPDMG | reducedStormData$CROPDMG)
filter(reducedStormData, damageNA)
```
No NA for property or crop damage.

```{r}
unique(reducedStormData$EVTYPE)
```

There are leading and tailing space in the storm type.

2. Clean up EVTYPE
Beside remove leading and trailing space, we also convert all strom type to upper case
```{r}
reducedStormData$EVTYPE <- trimws(reducedStormData$EVTYPE, which = c("both"), whitespace = "[ \t\r\n]")
reducedStormData$EVTYPE <- tolower(reducedStormData$EVTYPE)
reducedStormData$EVTYPE <- toupper(reducedStormData$EVTYPE)
```


```{r}
evtypeLst <- sort(unique(reducedStormData$EVTYPE))
evtypeLst
```

### Remove Unwanted Rows

```{r}
removeEvtypeCondition <- c("?")
cleanStormData <- subset(reducedStormData, EVTYPE != removeEvtypeCondition)
```


```{r}
wrongType <- c("ASTRONOMICAL HIGH TIDE", "AVALANCE")
correctType <- c("ASTRONOMICAL LOW TIDE", "AVALANCHE")

```



Convert PROPDMGEXP and CROPDMGEXP
```{r}
findCon <- c("?","+","-","k","K","m","M","b","B")
replacement <- c(0,1,0,1000,1000,1000000,1000000,1000000000,1000000000)
reducedData$PROPDMGEXP <- replace(reducedData$PROPDMGEXP, reducedData$PROPDMGEXP %in% findCon, replacement)
```

2. We then create a dataframe that is group by the event type, removing fatalities and injuries that are 0.
```{r}
fatalDf <- reducedData %>%
    group_by(EVTYPE) %>%
    filter(FATALITIES != 0 | INJURIES != 0) %>%
    summarise(FATALITIES_TOTAL=sum(FATALITIES), INJURIES_TOTAL=sum(INJURIES)) %>%
    arrange(desc(FATALITIES_TOTAL))

```

3. We select the top 10 event that has the worst fatalities and/or injuries
```{r}
rankedFatalities <- fatalDf[1:10,]
rankedFatalities
```

4. We need to convert and melt the dataframe into a format for plotting
```{r}
plotData <- as.data.frame(rankedFatalities)
meltedPlotData <- melt(plotData, id.vars="EVTYPE", variable_name = "Casualties_Type")
meltedPlotData
```

5. Plotting 
```{r}
ggplot(meltedPlotData, aes(fill=Casualties_Type,x=reorder(EVTYPE, -value),y=value))+ geom_bar(position="stack",stat="identity") + ylab("Number of Casualties")+ xlab("Type Of Disaster")+ggtitle("Top 10 Casualties Group by Disaster Type")+ theme(axis.text.x = element_text(angle=45, hjust=1))
```

processing data and convert the column
```{r}
fatalDf <- reducedData %>%
    group_by(EVTYPE) %>%
    filter(FATALITIES != 0 | INJURIES != 0) %>%
    summarise(FATALITIES_TOTAL=sum(FATALITIES), INJURIES_TOTAL=sum(INJURIES)) %>%
    arrange(desc(FATALITIES_TOTAL))

```

