---
title: "Reproducible Research Project Final"
output: html_document
date: "2024-02-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of Human and Economic Impact of Storms in United States

This report and the relevant data is available at <https://github.com/mlnotes2718/RepResearchFinalProject>

## SYNOPSIS

The following is an analysis of various kind of storm is the United States since 1950s. This analysis focus on human casualties and financial impact causes by various kind of storm. We need to answer the following questions:

1.  Across the United States, which types of events are most harmful with respect to population health?

2.  Across the United States, which types of events have the greatest economic consequences?

Based on the results, we concluded that the storm type that caused the most human casualties is **Tornado** and it is different from the storm type that has the most financial damage. The number one storm that caused the most financial damage is **Flood**. Hence policy maker should consider different type of response when addressing different type of

## DATA PROCESSING

### Setting Environment

```{r, message=FALSE}
# Import Library
library(dplyr)
library(stringr)
library(ggplot2)
library(reshape2)
```

### Data Source and Documentation

Data Source: [U.S. National Oceanic and Atmospheric Administration (NOAA) - Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

URL: <https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2>

Documentation of Data Source: [National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

[National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

### Download Data File

```{r, cache = TRUE}
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists("./StormData.csv.bz2")) {
    download.file(fileURL, destfile = "./StormData.csv.bz2")
}

```

### Data Dictionary

| Variable Name    | Description                                             |
|------------------|------------------------------------------------------|
| stormData        | The raw data that is read from StormData.csv.bz2        |
| reducedStormData | raw dataset with only only 7 columns                    |
| evtypeLst        | list that contains the unique event type of the dataset |
| cleanStormData   | cleaned dataset that have EVTYPE fixed.                 |
|                  |                                                         |

### Read File

```{r stormData,cache = TRUE, cache.extra = file.info("StormData.csv.bz2")}
stormData <- read.csv("StormData.csv.bz2")
```

File Structure

```{r}
str(stormData)
```

**Data contains 902297 with 37 columns.**

### Processing Data

1.  First we will filter out those data that has fatalities, injuries, property damage and crop damage.

```{r}
reducedStormData <- filter(stormData, FATALITIES != 0 | INJURIES !=0 |PROPDMG !=0 | CROPDMG != 0)
```

2.  We select relevant column for our analysis. We do not need all the 37 columns from the raw data.

```{r}
reducedStormData <- select(reducedStormData, STATE, BGN_DATE, EVTYPE, FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP,REMARKS)
```

Check for NA

```{r}
stormTypeNA <- is.na(reducedStormData$EVTYPE)
filter(reducedStormData, stormTypeNA)
```

No NA for storm type

```{r}
casualtiesNA <- is.na(reducedStormData$FATALITIES | reducedStormData$INJURIES)
filter(reducedStormData, casualtiesNA)
```

No NA for fatalities or injuries.

```{r}
damageNA <- is.na(reducedStormData$PROPDMG | reducedStormData$CROPDMG)
filter(reducedStormData, damageNA)
```

No NA for property or crop damage.

3.  File Structure after file size reduction.

```{r}
str(reducedStormData)
```

**We have 254633 rows with 10 columns.**

#### Examine Storm Event Classification

```{r}
unique(reducedStormData$EVTYPE)
```

-   There are 488 different type of storm classification. This does not match with the official 48 classification.
-   There are leading and tailing space in the storm type.
-   Some classification contains 2 space.
-   There is also a mixed of upper and lower case.

We need to clean up EVTYPE. Besides remove leading, trailing and double space, we also need to convert all storm type to upper case.

We also check the Property and Crop Damage Exponent

```{r}
unique(reducedStormData$PROPDMGEXP)
```

```{r}
unique(reducedStormData$CROPDMGEXP)
```

There is also mixed lower and upper which we can fixed together. The following code is to clean up trailing and leading space and upper and lower case for EVTYPE, PROPDMGEXP and CROPDMGEXP

```{r}
reducedStormData$EVTYPE <- trimws(reducedStormData$EVTYPE, which = c("both"), whitespace = "[ \t\r\n]")
reducedStormData$EVTYPE <- tolower(reducedStormData$EVTYPE)
reducedStormData$EVTYPE <- toupper(reducedStormData$EVTYPE)
reducedStormData$EVTYPE <- str_squish(reducedStormData$EVTYPE)

reducedStormData$PROPDMGEXP <- toupper(reducedStormData$PROPDMGEXP)
reducedStormData$CROPDMGEXP <- toupper(reducedStormData$CROPDMGEXP)
```

```{r}
evtypeLst <- sort(unique(reducedStormData$EVTYPE))
evtypeLst
```

**Misclassification reduce to 440**.

### Fixing Misclassification

The most ideal way to handle misclassification is to read the remarks for each event that are misclassified. However, due to constrain of resource, we adopt the strategy of reclassify those obvious spelling error by hand. We will not reclassify all misclassification. We are good as long as the unique classification is less than 100.

However, we did examine the remarks of 2 obvious misclassification. They are "?" and "APACHE COUNTY"

The following code is our clean up code.

For further details please refer to Appendix A in separate document.

```{r}
# Start Obvious Classification
reducedStormData$EVTYPE[grepl("ASTRONOMICAL HIGH TIDE", reducedStormData$EVTYPE)] <- "ASTRONOMICAL LOW TIDE"
reducedStormData$EVTYPE[grepl("AVALANCE", reducedStormData$EVTYPE)] <- "AVALANCHE"


# Blizzard
reducedStormData$EVTYPE[grepl("BLIZZARD/WINTER STORM", reducedStormData$EVTYPE)] <- "BLIZZARD"
reducedStormData$EVTYPE[grepl("GROUND BLIZZARD", reducedStormData$EVTYPE)] <- "BLIZZARD"


# Coastal Flood
reducedStormData$EVTYPE[grepl("^COASTAL*", reducedStormData$EVTYPE)] <- "COASTAL FLOOD"
reducedStormData$EVTYPE[grepl("EROSION/CSTL FLOOD", reducedStormData$EVTYPE)] <- "COASTAL FLOOD"


# COLD/WIND CHILL
reducedStormData$EVTYPE[grepl("^COLD*", reducedStormData$EVTYPE)] <- "COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("UNSEASONABLE COLD", reducedStormData$EVTYPE)] <- "COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("UNSEASONABLY COLD", reducedStormData$EVTYPE)] <- "COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("COOL AND WET", reducedStormData$EVTYPE)] <- "COLD/WIND CHILL"


# Dust Devil/Strom
reducedStormData$EVTYPE[grepl("^DUST DEVIL*", reducedStormData$EVTYPE)] <- "DUST DEVIL"
reducedStormData$EVTYPE[grepl("^DUST STORM*", reducedStormData$EVTYPE)] <- "DUST STORM"
reducedStormData$EVTYPE[grepl("^DRY MICROBURST*", reducedStormData$EVTYPE)] <- "DUST STORM"
reducedStormData$EVTYPE[grepl("BLOWING DUST", reducedStormData$EVTYPE)] <- "DUST STORM"


# Excessive Heat
reducedStormData$EVTYPE[grepl("^EXTREME HEAT*", reducedStormData$EVTYPE)] <- "EXCESSIVE HEAT"
reducedStormData$EVTYPE[grepl("^RECORD HEAT*", reducedStormData$EVTYPE)] <- "EXCESSIVE HEAT"
reducedStormData$EVTYPE[grepl("RECORD/EXCESSIVE HEAT", reducedStormData$EVTYPE)] <- "EXCESSIVE HEAT"


# Extreme Cold/Wind Chill
reducedStormData$EVTYPE[grepl("^EXTREME COLD*", reducedStormData$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("^EXTREME WIND*", reducedStormData$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("^EXTENDED*", reducedStormData$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("EXCESSIVE WETNESS", reducedStormData$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("RECORD COLD", reducedStormData$EVTYPE)] <- "EXTREME COLD/WIND CHILL"

# Flash Flood
reducedStormData$EVTYPE[grepl("^FLASH FLOOD*", reducedStormData$EVTYPE)] <- "FLASH FLOOD"

# Other Flood
reducedStormData$EVTYPE[grepl("^FLOOD*", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("^URBAN*", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("^RIVER.*FLOOD", reducedStormData$EVTYPE)] <- "FLOOD"

reducedStormData$EVTYPE[grepl("^MINOR FLOODING", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("MAJOR FLOOD", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("BREAKUP FLOODING", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("SMALL STREAM FLOOD", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("RURAL FLOOD", reducedStormData$EVTYPE)] <- "FLOOD"

# Freezing Fog
reducedStormData$EVTYPE[grepl("FOG", reducedStormData$EVTYPE)] <- "FREEZING FOG"

# Frost/Freeze
reducedStormData$EVTYPE[grepl("FROST\\FREEZE", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("FROST", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("FREEZE", reducedStormData$EVTYPE)] <- "FROST/FREEZE"

reducedStormData$EVTYPE[grepl("FREEZING RAIN", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("FREEZING DRIZZLE", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("FREEZING SPRAY", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("^HYPOTHERMIA.*", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("HYPERTHERMIA/EXPOSURE", reducedStormData$EVTYPE)] <- "FROST/FREEZE"


# Hail
reducedStormData$EVTYPE[grepl("^HAIL*", reducedStormData$EVTYPE)] <- "HAIL"


# Heat
reducedStormData$EVTYPE[grepl("^HEAT *", reducedStormData$EVTYPE)] <- "HEAT"


# Heavy Rain
reducedStormData$EVTYPE[grepl("^RAIN*", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("^HEAVY RAIN*", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("HEAVY PRECIPITATION", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("HEAVY SHOWER", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("EXCESSIVE RAINFALL", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("RECORD RAINFALL", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("UNSEASONAL RAIN", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("HVY RAIN", reducedStormData$EVTYPE)] <- "HEAVY RAIN"


# Heavy Snow
reducedStormData$EVTYPE[grepl("^HEAVY SNOW*", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("BLOWING SNOW", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("FALLING SNOW/ICE", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("LATE SEASON SNOW", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("EXCESSIVE SNOW", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("HEAVY LAKE SNOW", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("RECORD SNOW", reducedStormData$EVTYPE)] <- "HEAVY SNOW"


# High Surf
reducedStormData$EVTYPE[grepl("^HEAVY SURF*", reducedStormData$EVTYPE)] <- "HIGH SURF"
reducedStormData$EVTYPE[grepl("HAZARDOUS SURF", reducedStormData$EVTYPE)] <- "HIGH SURF"
reducedStormData$EVTYPE[grepl("HIGH SURF ADVISORY", reducedStormData$EVTYPE)] <- "HIGH SURF"
reducedStormData$EVTYPE[grepl("HIGH SWELLS", reducedStormData$EVTYPE)] <- "HIGH SURF"
reducedStormData$EVTYPE[grepl("HIGH WAVES", reducedStormData$EVTYPE)] <- "HIGH SURF"


# High Wind
reducedStormData$EVTYPE[grepl("^HIGH WIND*", reducedStormData$EVTYPE)] <- "HIGH WIND"


# Hurricane (Typhoon)
reducedStormData$EVTYPE[grepl("^HURRICANE*", reducedStormData$EVTYPE)] <- "HURRICANE (TYPHOON)"
reducedStormData$EVTYPE[grepl("^TYPHOON", reducedStormData$EVTYPE)] <- "HURRICANE (TYPHOON)"


# Ice Storm
reducedStormData$EVTYPE[grepl("^ICE*", reducedStormData$EVTYPE)] <- "ICE STORM"
reducedStormData$EVTYPE[grepl("BLACK ICE", reducedStormData$EVTYPE)] <- "ICE STORM"
reducedStormData$EVTYPE[grepl("^GLAZE", reducedStormData$EVTYPE)] <- "ICE STORM"


# Lakeshore Flood
reducedStormData$EVTYPE[grepl("^LAKE FLOOD", reducedStormData$EVTYPE)] <- "LAKESHORE FLOOD"


# Lake-Effect Snow
reducedStormData$EVTYPE[grepl("^LAKE EFFECT SNOW", reducedStormData$EVTYPE)] <- "LAKE-EFFECT SNOW"


# Lightning
reducedStormData$EVTYPE[grepl("^LIGHT*", reducedStormData$EVTYPE)] <- "LIGHTNING"
reducedStormData$EVTYPE[grepl("^LIGNTNING", reducedStormData$EVTYPE)] <- "LIGHTNING"


# Marine Thunderstorm Wind
reducedStormData$EVTYPE[grepl("^MARINE TSTM WIND", reducedStormData$EVTYPE)] <- "MARINE THUNDERSTORM WIND"


# Rip Current
reducedStormData$EVTYPE[grepl("^RIP CURRENT.*", reducedStormData$EVTYPE)] <- "RIP CURRENT"


# Strong Wind 
reducedStormData$EVTYPE[grepl("^GUST*", reducedStormData$EVTYPE)] <- "STRONG WIND"
reducedStormData$EVTYPE[grepl("^STRONG WIND*", reducedStormData$EVTYPE)] <- "STRONG WIND"
reducedStormData$EVTYPE[grepl("^WIND", reducedStormData$EVTYPE)] <- "STRONG WIND"


# Thunderstorm Wind
reducedStormData$EVTYPE[grepl("^TSTM*", reducedStormData$EVTYPE)] <- "THUNDERSTORM WIND"
reducedStormData$EVTYPE[grepl("^THU*", reducedStormData$EVTYPE)] <- "THUNDERSTORM WIND"
reducedStormData$EVTYPE[grepl("^TUNDERSTORM*", reducedStormData$EVTYPE)] <- "THUNDERSTORM WIND"
reducedStormData$EVTYPE[grepl("^SEVERE*", reducedStormData$EVTYPE)] <- "THUNDERSTORM WIND"


# Tornado
reducedStormData$EVTYPE[grepl("^TORNADO*", reducedStormData$EVTYPE)] <- "TORNADO"
reducedStormData$EVTYPE[grepl("^TORNDAO*", reducedStormData$EVTYPE)] <- "TORNADO"


# Tropical Storm
reducedStormData$EVTYPE[grepl("^TROPICAL STORM*", reducedStormData$EVTYPE)] <- "TROPICAL STORM"


# Waterspout
reducedStormData$EVTYPE[grepl("^WATERSPOUT*", reducedStormData$EVTYPE)] <- "WATERSPOUT"


# Winter Storm
reducedStormData$EVTYPE[grepl("^WINTER STORM*", reducedStormData$EVTYPE)] <- "WINTER STORM"


# Winter Weather
reducedStormData$EVTYPE[grepl("^WINTER WEATHER*", reducedStormData$EVTYPE)] <- "WINTER WEATHER"
reducedStormData$EVTYPE[grepl("WINTRY MIX", reducedStormData$EVTYPE)] <- "WINTER WEATHER"
reducedStormData$EVTYPE[grepl("^SNOW.*", reducedStormData$EVTYPE)] <- "WINTER WEATHER"



# Wildfire
reducedStormData$EVTYPE[grepl("^WILD.*FIRE*", reducedStormData$EVTYPE)] <- "WILDFIRE"
reducedStormData$EVTYPE[grepl("^BRUSH FIRE", reducedStormData$EVTYPE)] <- "WILDFIRE"
reducedStormData$EVTYPE[grepl("^FOREST FIRE*", reducedStormData$EVTYPE)] <- "WILDFIRE"
reducedStormData$EVTYPE[grepl("^GRASS FIRE*", reducedStormData$EVTYPE)] <- "WILDFIRE"


# Re-classify APACHE COUNTY
reducedStormData$EVTYPE[grepl("APACHE COUNTY", reducedStormData$EVTYPE)] <- "THUNDERSTORM WIND"
```

### Remove Unwanted Rows

```{r}
removeEvtypeCondition <- c("?")
cleanStormData <- subset(reducedStormData, EVTYPE != removeEvtypeCondition)
```

File Structure of Clean Data

```{r}
str(cleanStormData)
```

**We have a total of 254632 rows and 10 columns.**

### Convert PROPDMGEXP and CROPDMGEXP

We also need to convert the exponent of the property and crop damage. The exponent of k or K represent thousands. M or m in millions. Any number is equivalent to 10 to the power of the number. (e.g. 7 = 10\^7)

First we need to find out the unique exponent of PROPDMGEXP and CROPDMGEXP.

```{r}
uniquePROPDMGEXP <- unique(cleanStormData$PROPDMGEXP)
uniqueCROPDMGEXP <- unique(cleanStormData$CROPDMGEXP)
uniqueBothDMGEXP <- union(uniquePROPDMGEXP, uniqueCROPDMGEXP)
uniqueBothDMGEXP <- sort(uniqueBothDMGEXP)
uniqueBothDMGEXP
```

The empty string "" will caused a problem, so we will replace those empty string with '-' on the clean dataset.

```{r}
cleanStormData$PROPDMGEXP[cleanStormData$PROPDMGEXP == ""] <- '-'
cleanStormData$CROPDMGEXP[cleanStormData$CROPDMGEXP == ""] <- '-'
```

Next, we regenerate the unique values again

```{r}
uniquePROPDMGEXP <- unique(cleanStormData$PROPDMGEXP)
uniqueCROPDMGEXP <- unique(cleanStormData$CROPDMGEXP)
uniqueBothDMGEXP <- union(uniquePROPDMGEXP, uniqueCROPDMGEXP)
uniqueBothDMGEXP <- sort(uniqueBothDMGEXP)
uniqueBothDMGEXP
```

The corresponding replacement values will be as follows:

```{r}
replacement <- c(0,0,1,10^0,10^2,10^3,10^4,10^5,10^6,10^7,1000000000,100,1000,1000000)
names(replacement) <- uniqueBothDMGEXP
replacement
```

We create a new column called PROPDMGEXPNUM which store the numeric values based on the reference above. Similar for CROPDMGEXPNUM. We also compute the total damage in new column PROPDMGSUM and CROPDMGSUM

```{r}
cleanStormData <- cleanStormData %>%
  mutate(PROPDMGEXPNUM = replacement[cleanStormData$PROPDMGEXP]) %>%
  mutate(CROPDMGEXPNUM = replacement[cleanStormData$CROPDMGEXP]) %>%
  mutate(PROPDMGSUM = PROPDMG * PROPDMGEXPNUM) %>%
  mutate(CROPDMGSUM = CROPDMG * CROPDMGEXPNUM) 
```

We use select function to re-order the columns

```{r}
cleanStormData <- select(cleanStormData, STATE, BGN_DATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, PROPDMGEXPNUM, PROPDMGSUM, CROPDMG, CROPDMGEXP,CROPDMGSUM, REMARKS)
head(cleanStormData)
```

Next, We then create a data frame that is group by the event type, removing fatalities and injuries that are 0.

```{r}
fatalDf <- cleanStormData %>%
    group_by(EVTYPE) %>%
    filter(FATALITIES != 0 | INJURIES != 0) %>%
    summarise(FATALITIES_TOTAL=sum(FATALITIES), INJURIES_TOTAL=sum(INJURIES)) %>%
    arrange(desc(FATALITIES_TOTAL))

```

We select the top 10 event that has the worst fatalities and/or injuries

```{r}
rankedFatalities <- fatalDf[1:10,]
rankedFatalities
```

We need to convert and melt the dataframe into a format for plotting

```{r}
plotData <- as.data.frame(rankedFatalities)
meltedPlotData <- melt(plotData, id.vars="EVTYPE", variable.name = "Casualties")
meltedPlotData
```

```{r}
econDf <- cleanStormData %>%
    group_by(EVTYPE) %>%
    filter(PROPDMGSUM != 0 | CROPDMGSUM != 0) %>%
    summarise(PROPDMG_TOTAL=sum(PROPDMGSUM), CROPDMG_TOTAL=sum(CROPDMGSUM)) %>%
    arrange(desc(PROPDMG_TOTAL))
```

```{r}
econDf[1:10,]
```

```{r}
rankedEconomicImpact <- econDf[1:10,]
rankedEconomicImpact$PROPDMG_TOTAL <- rankedEconomicImpact$PROPDMG_TOTAL/1000000000
rankedEconomicImpact$CROPDMG_TOTAL <- rankedEconomicImpact$CROPDMG_TOTAL/1000000000
rankedEconomicImpact <- rankedEconomicImpact %>%
  rename(Property_Damage=PROPDMG_TOTAL, Crop_Damage = CROPDMG_TOTAL)
rankedEconomicImpact
```

```{r}
plotDataEcon <- as.data.frame(rankedEconomicImpact)
meltedPlotDataEcon <- melt(plotDataEcon, id.vars="EVTYPE", variable.name = "Damages")
meltedPlotDataEcon
```

## Results

```{r}
ggplot(meltedPlotData, aes(fill=Casualties,x=reorder(EVTYPE, value),y=value))+ geom_bar(position="stack",stat="identity") + ylab("Number of Casualties")+ xlab("Type Of Disaster")+ggtitle("Top 10 Casualties Group by Disaster Type")+ coord_flip()+ facet_grid(. ~ Casualties)+
theme(legend.position="bottom")
```

```{r}
rankedFatalities
```

```{r}
ggplot(meltedPlotDataEcon, aes(fill=Damages,x=reorder(EVTYPE, value),y=value))+ geom_bar(position="stack",stat="identity") + ylab("Total Economic Damages in Billions")+ xlab("Type Of Disaster")+ggtitle("Top 10 Economic damages Group by Disaster Type")+ coord_flip()+ 
facet_grid(. ~ Damages)+
theme(legend.position="bottom")
```

```{r}
rankedEconomicImpact
```

## Conclusion

Based on the results above, policy maker can rank the storm type that caused the most fatalities and perform the same for those storm type that caused the most economic damage.
