---
title: "Reproducible Research Project Final"
output: html_document
date: "2024-02-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of Human and Economic Impact of Storms in United States
This report and teh relevant data is available at [https://github.com/mlnotes2718/RepResearchFinalProject](https://github.com/mlnotes2718/RepResearchFinalProject)

## SYNOPSIS
The following is an analysis of various kind of storm is the United States since 1950s. This analysis focus on human casualties and financial impact causes by various kind of storm.


## DATA PROCESSING

### Setting Envionment
```{r, message=FALSE}
# Import Library
library(dplyr)
library(stringr)
library(stringdist)
library(ggplot2)
library(reshape2)
```

### Data Source and Documentation

Data Source: [U.S. National Oceanic and Atmospheric Administration (NOAA) - Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

URL: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

Documentation of Data Source: 
[National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

[National Climatic Data Center Storm Events FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)



### Download Data File
```{r, cache = TRUE}
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileURL, destfile = "./StormData.csv.bz2")
```

### Data Dictionary
|Variable Name | Description |
|---|---|
|stormData | The raw data that is read from StormData.csv.bz2|
|reducedData | raw dataset with only only 7 columns |


### Read File
```{r, cache=TRUE}
stormData <- read.csv("StormData.csv.bz2")
```

File Structure
```{r}
str(stormData)
```

**Data contains 902297 with 37 columns.**

### Processing Data
1. First we will filter out those data that has fatalities, injuries, property damage and crop damage. 
```{r}
reducedStormData <- filter(stormData, FATALITIES != 0 | INJURIES !=0 |PROPDMG !=0 | CROPDMG != 0)
```

2. Select relevant column
```{r}
reducedStormData <- select(reducedStormData, STATE, BGN_DATE, EVTYPE, FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP,REMARKS)
```

Check for NA
```{r}
stormTypeNA <- is.na(reducedStormData$EVTYPE)
filter(reducedStormData, stormTypeNA)
```
No NA for storm type


```{r}
casualtiesNA <- is.na(reducedStormData$FATALITIES | reducedStormData$INJURIES)
filter(reducedStormData, casualtiesNA)
```

No NA for fatalities or injuries.

```{r}
damageNA <- is.na(reducedStormData$PROPDMG | reducedStormData$CROPDMG)
filter(reducedStormData, damageNA)
```
No NA for property or crop damage.

#### Examine Storm Event Classification
```{r}
unique(reducedStormData$EVTYPE)
```

- There are 488 different type of storm classification. This does not match with the official 48 classification. 
- There are leading and tailing space in the storm type.
- Some classification contains 2 space.
- There is also a mixed of upper and lower case.

Clean up EVTYPE
Beside remove leading, trailing and double space, we also convert all storm type to upper case.
```{r}
reducedStormData$EVTYPE <- trimws(reducedStormData$EVTYPE, which = c("both"), whitespace = "[ \t\r\n]")
reducedStormData$EVTYPE <- tolower(reducedStormData$EVTYPE)
reducedStormData$EVTYPE <- toupper(reducedStormData$EVTYPE)
reducedStormData$EVTYPE <- str_squish(reducedStormData$EVTYPE)
```


```{r}
evtypeLst <- sort(unique(reducedStormData$EVTYPE))
evtypeLst
```
**Misclassification reduce to 440**.

### Fixing Misclassification
The most ideal way to handle misclassification is to read the remarks for each event that are misclassified. However, due to constrain of resource, we adopt the strategy of reclassify those obvious spelling error by hand. We will not reclassify all misclassification. We are good as long as the unqiue classification is less than 100.

However, we did examine the remarks details of 2 obvious misclassification. They are "?" and "APACHE COUNTY"

The following code is our clean up code.

For further details please refer to Appendix A in separarte document.
```{r}
# Start Obvious Classification
reducedStormData$EVTYPE[grepl("ASTRONOMICAL HIGH TIDE", reducedStormData$EVTYPE)] <- "ASTRONOMICAL LOW TIDE"
reducedStormData$EVTYPE[grepl("AVALANCE", reducedStormData$EVTYPE)] <- "AVALANCHE"


# Blizzard
reducedStormData$EVTYPE[grepl("BLIZZARD/WINTER STORM", reducedStormData$EVTYPE)] <- "BLIZZARD"
reducedStormData$EVTYPE[grepl("GROUND BLIZZARD", reducedStormData$EVTYPE)] <- "BLIZZARD"


# Coastal Flood
reducedStormData$EVTYPE[grepl("^COASTAL*", reducedStormData$EVTYPE)] <- "COASTAL FLOOD"
reducedStormData$EVTYPE[grepl("EROSION/CSTL FLOOD", reducedStormData$EVTYPE)] <- "COASTAL FLOOD"


# COLD/WIND CHILL
reducedStormData$EVTYPE[grepl("^COLD*", reducedStormData$EVTYPE)] <- "COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("UNSEASONABLE COLD", reducedStormData$EVTYPE)] <- "COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("UNSEASONABLY COLD", reducedStormData$EVTYPE)] <- "COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("COOL AND WET", reducedStormData$EVTYPE)] <- "COLD/WIND CHILL"


# Dust Devil/Strom
reducedStormData$EVTYPE[grepl("^DUST DEVIL*", reducedStormData$EVTYPE)] <- "DUST DEVIL"
reducedStormData$EVTYPE[grepl("^DUST STORM*", reducedStormData$EVTYPE)] <- "DUST STORM"
reducedStormData$EVTYPE[grepl("^DRY MICROBURST*", reducedStormData$EVTYPE)] <- "DUST STORM"
reducedStormData$EVTYPE[grepl("BLOWING DUST", reducedStormData$EVTYPE)] <- "DUST STORM"


# Excessive Heat
reducedStormData$EVTYPE[grepl("^EXTREME HEAT*", reducedStormData$EVTYPE)] <- "EXCESSIVE HEAT"
reducedStormData$EVTYPE[grepl("^RECORD HEAT*", reducedStormData$EVTYPE)] <- "EXCESSIVE HEAT"
reducedStormData$EVTYPE[grepl("RECORD/EXCESSIVE HEAT", reducedStormData$EVTYPE)] <- "EXCESSIVE HEAT"


# Extreme Cold/Wind Chill
reducedStormData$EVTYPE[grepl("^EXTREME COLD*", reducedStormData$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("^EXTREME WIND*", reducedStormData$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("^EXTENDED*", reducedStormData$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("EXCESSIVE WETNESS", reducedStormData$EVTYPE)] <- "EXTREME COLD/WIND CHILL"
reducedStormData$EVTYPE[grepl("RECORD COLD", reducedStormData$EVTYPE)] <- "EXTREME COLD/WIND CHILL"

# Flash Flood
reducedStormData$EVTYPE[grepl("^FLASH FLOOD*", reducedStormData$EVTYPE)] <- "FLASH FLOOD"

# Other Flood
reducedStormData$EVTYPE[grepl("^FLOOD*", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("^URBAN*", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("^RIVER.*FLOOD", reducedStormData$EVTYPE)] <- "FLOOD"

reducedStormData$EVTYPE[grepl("^MINOR FLOODING", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("MAJOR FLOOD", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("BREAKUP FLOODING", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("SMALL STREAM FLOOD", reducedStormData$EVTYPE)] <- "FLOOD"
reducedStormData$EVTYPE[grepl("RURAL FLOOD", reducedStormData$EVTYPE)] <- "FLOOD"

# Freezing Fog
reducedStormData$EVTYPE[grepl("FOG", reducedStormData$EVTYPE)] <- "FREEZING FOG"

# Frost/Freeze
reducedStormData$EVTYPE[grepl("FROST\\FREEZE", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("FROST", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("FREEZE", reducedStormData$EVTYPE)] <- "FROST/FREEZE"

reducedStormData$EVTYPE[grepl("FREEZING RAIN", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("FREEZING DRIZZLE", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("FREEZING SPRAY", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("^HYPOTHERMIA.*", reducedStormData$EVTYPE)] <- "FROST/FREEZE"
reducedStormData$EVTYPE[grepl("HYPERTHERMIA/EXPOSURE", reducedStormData$EVTYPE)] <- "FROST/FREEZE"


# Hail
reducedStormData$EVTYPE[grepl("^HAIL*", reducedStormData$EVTYPE)] <- "HAIL"


# Heat
reducedStormData$EVTYPE[grepl("^HEAT *", reducedStormData$EVTYPE)] <- "HEAT"


# Heavy Rain
reducedStormData$EVTYPE[grepl("^RAIN*", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("^HEAVY RAIN*", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("HEAVY PRECIPITATION", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("HEAVY SHOWER", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("EXCESSIVE RAINFALL", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("RECORD RAINFALL", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("UNSEASONAL RAIN", reducedStormData$EVTYPE)] <- "HEAVY RAIN"
reducedStormData$EVTYPE[grepl("HVY RAIN", reducedStormData$EVTYPE)] <- "HEAVY RAIN"


# Heavy Snow
reducedStormData$EVTYPE[grepl("^HEAVY SNOW*", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("BLOWING SNOW", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("FALLING SNOW/ICE", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("LATE SEASON SNOW", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("EXCESSIVE SNOW", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("HEAVY LAKE SNOW", reducedStormData$EVTYPE)] <- "HEAVY SNOW"
reducedStormData$EVTYPE[grepl("RECORD SNOW", reducedStormData$EVTYPE)] <- "HEAVY SNOW"


# High Surf
reducedStormData$EVTYPE[grepl("^HEAVY SURF*", reducedStormData$EVTYPE)] <- "HIGH SURF"
reducedStormData$EVTYPE[grepl("HAZARDOUS SURF", reducedStormData$EVTYPE)] <- "HIGH SURF"
reducedStormData$EVTYPE[grepl("HIGH SURF ADVISORY", reducedStormData$EVTYPE)] <- "HIGH SURF"
reducedStormData$EVTYPE[grepl("HIGH SWELLS", reducedStormData$EVTYPE)] <- "HIGH SURF"
reducedStormData$EVTYPE[grepl("HIGH WAVES", reducedStormData$EVTYPE)] <- "HIGH SURF"


# High Wind
reducedStormData$EVTYPE[grepl("^HIGH WIND*", reducedStormData$EVTYPE)] <- "HIGH WIND"


# Hurricane (Typhoon)
reducedStormData$EVTYPE[grepl("^HURRICANE*", reducedStormData$EVTYPE)] <- "HURRICANE (TYPHOON)"
reducedStormData$EVTYPE[grepl("^TYPHOON", reducedStormData$EVTYPE)] <- "HURRICANE (TYPHOON)"


# Ice Storm
reducedStormData$EVTYPE[grepl("^ICE*", reducedStormData$EVTYPE)] <- "ICE STORM"
reducedStormData$EVTYPE[grepl("BLACK ICE", reducedStormData$EVTYPE)] <- "ICE STORM"
reducedStormData$EVTYPE[grepl("^GLAZE", reducedStormData$EVTYPE)] <- "ICE STORM"


# Lakeshore Flood
reducedStormData$EVTYPE[grepl("^LAKE FLOOD", reducedStormData$EVTYPE)] <- "LAKESHORE FLOOD"


# Lake-Effect Snow
reducedStormData$EVTYPE[grepl("^LAKE EFFECT SNOW", reducedStormData$EVTYPE)] <- "LAKE-EFFECT SNOW"


# Lightning
reducedStormData$EVTYPE[grepl("^LIGHT*", reducedStormData$EVTYPE)] <- "LIGHTNING"
reducedStormData$EVTYPE[grepl("^LIGNTNING", reducedStormData$EVTYPE)] <- "LIGHTNING"


# Marine Thunderstorm Wind
reducedStormData$EVTYPE[grepl("^MARINE TSTM WIND", reducedStormData$EVTYPE)] <- "MARINE THUNDERSTORM WIND"


# Rip Current
reducedStormData$EVTYPE[grepl("^RIP CURRENT.*", reducedStormData$EVTYPE)] <- "RIP CURRENT"


# Strong Wind 
reducedStormData$EVTYPE[grepl("^GUST*", reducedStormData$EVTYPE)] <- "STRONG WIND"
reducedStormData$EVTYPE[grepl("^STRONG WIND*", reducedStormData$EVTYPE)] <- "STRONG WIND"
reducedStormData$EVTYPE[grepl("^WIND", reducedStormData$EVTYPE)] <- "STRONG WIND"


# Thunderstorm Wind
reducedStormData$EVTYPE[grepl("^TSTM*", reducedStormData$EVTYPE)] <- "THUNDERSTORM WIND"
reducedStormData$EVTYPE[grepl("^THU*", reducedStormData$EVTYPE)] <- "THUNDERSTORM WIND"
reducedStormData$EVTYPE[grepl("^TUNDERSTORM*", reducedStormData$EVTYPE)] <- "THUNDERSTORM WIND"
reducedStormData$EVTYPE[grepl("^SEVERE*", reducedStormData$EVTYPE)] <- "THUNDERSTORM WIND"


# Tornado
reducedStormData$EVTYPE[grepl("^TORNADO*", reducedStormData$EVTYPE)] <- "TORNADO"
reducedStormData$EVTYPE[grepl("^TORNDAO*", reducedStormData$EVTYPE)] <- "TORNADO"


# Tropical Storm
reducedStormData$EVTYPE[grepl("^TROPICAL STORM*", reducedStormData$EVTYPE)] <- "TROPICAL STORM"


# Waterspout
reducedStormData$EVTYPE[grepl("^WATERSPOUT*", reducedStormData$EVTYPE)] <- "WATERSPOUT"


# Winter Storm
reducedStormData$EVTYPE[grepl("^WINTER STORM*", reducedStormData$EVTYPE)] <- "WINTER STORM"


# Winter Weather
reducedStormData$EVTYPE[grepl("^WINTER WEATHER*", reducedStormData$EVTYPE)] <- "WINTER WEATHER"
reducedStormData$EVTYPE[grepl("WINTRY MIX", reducedStormData$EVTYPE)] <- "WINTER WEATHER"
reducedStormData$EVTYPE[grepl("^SNOW.*", reducedStormData$EVTYPE)] <- "WINTER WEATHER"



# Wildfire
reducedStormData$EVTYPE[grepl("^WILD.*FIRE*", reducedStormData$EVTYPE)] <- "WILDFIRE"
reducedStormData$EVTYPE[grepl("^BRUSH FIRE", reducedStormData$EVTYPE)] <- "WILDFIRE"
reducedStormData$EVTYPE[grepl("^FOREST FIRE*", reducedStormData$EVTYPE)] <- "WILDFIRE"
reducedStormData$EVTYPE[grepl("^GRASS FIRE*", reducedStormData$EVTYPE)] <- "WILDFIRE"


# Re-classify APACHE COUNTY
reducedStormData$EVTYPE[grepl("APACHE COUNTY", reducedStormData$EVTYPE)] <- "THUNDERSTORM WIND"
```



### Remove Unwanted Rows

```{r}
removeEvtypeCondition <- c("?")
cleanStormData <- subset(reducedStormData, EVTYPE != removeEvtypeCondition)
```





Convert PROPDMGEXP and CROPDMGEXP
```{r}
findCon <- c("?","+","-","k","K","m","M","b","B")
replacement <- c(0,1,0,1000,1000,1000000,1000000,1000000000,1000000000)
reducedStormData$PROPDMGEXP <- replace(reducedStormData$PROPDMGEXP, reducedStormData$PROPDMGEXP %in% findCon, replacement)
```

2. We then create a dataframe that is group by the event type, removing fatalities and injuries that are 0.
```{r}
fatalDf <- reducedStormData %>%
    group_by(EVTYPE) %>%
    filter(FATALITIES != 0 | INJURIES != 0) %>%
    summarise(FATALITIES_TOTAL=sum(FATALITIES), INJURIES_TOTAL=sum(INJURIES)) %>%
    arrange(desc(FATALITIES_TOTAL))

```

3. We select the top 10 event that has the worst fatalities and/or injuries
```{r}
rankedFatalities <- fatalDf[1:10,]
rankedFatalities
```

4. We need to convert and melt the dataframe into a format for plotting
```{r}
plotData <- as.data.frame(rankedFatalities)
meltedPlotData <- melt(plotData, id.vars="EVTYPE")
meltedPlotData
```

5. Plotting 
```{r}
ggplot(meltedPlotData, aes(fill=variable,x=reorder(EVTYPE, -value),y=value))+ geom_bar(position="stack",stat="identity") + ylab("Number of Casualties")+ xlab("Type Of Disaster")+ggtitle("Top 10 Casualties Group by Disaster Type")+ theme(axis.text.x = element_text(angle=45, hjust=1))
```

processing data and convert the column
```{r}
fatalDf <- reducedData %>%
    group_by(EVTYPE) %>%
    filter(FATALITIES != 0 | INJURIES != 0) %>%
    summarise(FATALITIES_TOTAL=sum(FATALITIES), INJURIES_TOTAL=sum(INJURIES)) %>%
    arrange(desc(FATALITIES_TOTAL))

```

